---
title: "Modeling Productivity in an Aging Population"
description: "Interactive exploration of how demographic changes and technological progress affect economic productivity"
date: 2025-09-26
categories: [Economics, Demographics, Bayesian]
tags: ["productivity", "aging", "demographics", "bayesian", "brms", "interactive"]
draft: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: true
format:
  html:
    embed-resources: false
    self-contained: false
    minimal: true
    code-fold: false
    code-tools: false
    title-block-style: none
    toc: false
    theme: none
prefer-html: true
knitr:
  opts_chunk:
    fig.width: 8
    fig.height: 5
    dpi: 150
---

```{r setup}
#| include: false

library(tidyverse)
library(here)
library(plotly)
library(DT)
library(brms)
library(tidybayes)
library(readr)
library(lubridate)

# Set theme for plots
theme_set(theme_minimal(base_size = 12))

# Data directory already exists
```

## Introduction

[Your introduction text here]

## Literature Review

[Your literature review here]

## Data and Methodology

### Data Sources

```{r}
#| label: data-loading
#| message: false

# Load all helper functions
source(here("src", "content", "blog", "2025", "09-26-modeling-productivity-aging-population", "R", "data_loading.R"))
source(here("src", "content", "blog", "2025", "09-26-modeling-productivity-aging-population", "R", "visualization_helpers.R"))
source(here("src", "content", "blog", "2025", "09-26-modeling-productivity-aging-population", "R", "bayesian_modeling.R"))
source(here("src", "content", "blog", "2025", "09-26-modeling-productivity-aging-population", "R", "productivity_index.R"))
source(here("src", "content", "blog", "2025", "09-26-modeling-productivity-aging-population", "R", "stationarity_analysis.R"))
source(here("src", "content", "blog", "2025", "09-26-modeling-productivity-aging-population", "R", "model_validation.R"))

# Load the datasets
germany_productivity <- load_productivity_data()
germany_demographics <- load_demographic_data()
cognitive_curves <- load_cognitive_data()
```

### Effective Productivity Index (EPI)

**Addressing the Model Improvement Feedback:**

Based on analysis feedback, the original model used crude demographic aggregates (weighted age, population shares) that couldn't capture the non-linear age-productivity relationship. We now implement an **Effective Productivity Index (EPI)** that properly weights each age group by their cognitive productivity.

```{r}
#| label: epi-calculation

# Calculate EPI using different methods
epi_data <- calculate_epi(germany_demographics, cognitive_curves, method = "composite")

# Compare EPI calculation methods
epi_comparison <- compare_epi_methods(germany_demographics, cognitive_curves)

# Display EPI time series
epi_plots <- create_epi_plots(epi_data, epi_comparison)
epi_plots$epi_timeseries
```

### Stationarity Analysis

Another key issue was the AR(1) coefficient ≈ 1.05 that dominated demographic signals. We test for unit roots and assess whether to model levels or growth rates.

```{r}
#| label: stationarity-tests

# Test stationarity of productivity series
productivity_stationarity <- test_stationarity(germany_productivity, "productivity_per_hour")

# Test stationarity of EPI
epi_stationarity <- test_stationarity(epi_data, "epi_normalized")

# Create stationarity diagnostic plots
stationarity_plots <- create_stationarity_plots(germany_productivity, "productivity_per_hour")
stationarity_plots$timeseries
stationarity_plots$first_differences
```

### Model Specification

Based on stationarity results, we implement multiple model specifications:

1. **EPI Levels Model**: Uses EPI instead of crude demographics
2. **EPI Growth Model**: Models growth rates if non-stationary
3. **Original Model**: For comparison

```{r}
#| label: model-specification

# Prepare data for multiple model types
model_data <- prepare_model_data(germany_productivity, epi_data, stationarity_results, ensure_consistent_sample = TRUE)

# Display data preparation summary
cat("Model data spans", min(model_data$year), "to", max(model_data$year), "\n")
cat("Key variables: EPI, log productivity, growth rates\n")
cat("Sample size:", nrow(model_data), "observations\n")

head(model_data %>% select(year, log_productivity, epi_normalized, productivity_growth, epi_growth))
```

## Interactive Model Exploration

### Age-Competency Relationship

```{r}
#| label: age-competency-viz

# Create interactive age-competency curve using helper function
p_age_comp <- create_age_competency_plot(cognitive_curves)
p_age_comp
```

### Technology Progress Impact

```{r}
#| label: tech-progress-viz

# Create technology progress scenarios using helper function
p_tech <- create_tech_progress_plot(start_year = 1970, end_year = 2050)
p_tech
```

### Demographic Evolution

```{r}
#| label: demographics-viz

# Create demographic evolution plot using helper function
p_demo <- create_demographic_plot(germany_demographics)
p_demo
```

## Improved Bayesian Model Fitting

### Model Comparison

We fit multiple model specifications to address the identified issues:

```{r}
#| label: model-comparison
#| cache: true

# Fit multiple model specifications with improved error handling
model_types <- c("original", "epi_levels", "epi_growth")

# Fit models with error handling
tryCatch({
  models_list <- fit_model_comparison(model_data, model_types)

  # Perform model selection with fixed comparison function
  selection_results <- perform_model_selection(models_list, model_data)

  cat("✓ Model comparison completed successfully\n")
  cat("Best model:", selection_results$best_model_name, "\n")

  # Display results
  if (!is.null(selection_results$comparison_results$summary)) {
    print(selection_results$comparison_results$summary)
  }

  # Report sample size consistency
  if (selection_results$sample_size_consistent) {
    cat("\n✓ All models fitted on same sample size - valid comparison\n")
  } else {
    cat("\n⚠ Models have different sample sizes - see individual metrics\n")
  }

}, error = function(e) {
  cat("✗ Error in model comparison:\n")
  cat("Error:", e$message, "\n")

  # Fallback: fit just levels models that should work
  cat("\nTrying fallback with just levels models...\n")
  fallback_types <- c("original", "epi_levels")
  models_list <- fit_model_comparison(model_data, fallback_types)
  selection_results <- perform_model_selection(models_list, model_data)

  cat("Fallback best model:", selection_results$best_model_name, "\n")
})
```

### Enhanced Parameter Analysis

```{r}
#| label: parameter-analysis

# Extract parameters for best model (if available)
if (exists("selection_results") && !is.null(selection_results$best_model)) {
  # Extract parameters for best model
  best_model <- selection_results$best_model
  best_model_name <- selection_results$best_model_name

  # Get detailed parameter estimates
  parameter_results <- extract_parameter_estimates(best_model, best_model_name)

  # Display parameter summary
  parameter_results$summary %>%
    select(parameter, mean, median, q025, q975, significant) %>%
    DT::datatable(
      caption = "Parameter Estimates for Best Model",
      options = list(pageLength = 10, scrollX = TRUE)
    ) %>%
    DT::formatRound(columns = 2:5, digits = 4)
} else {
  cat("Parameter analysis not available - model comparison failed\n")
  cat("This may be due to computational issues or insufficient data\n")
}
```

### Enhanced Model Diagnostics

```{r}
#| label: enhanced-diagnostics

# Create comprehensive diagnostics for all models (if available)
if (exists("models_list") && exists("selection_results") && !is.null(selection_results$best_model)) {
  # Create comprehensive diagnostics for all models
  diagnostics_results <- create_enhanced_diagnostics(models_list, model_data)

  # Display diagnostics for best model
  best_model_name <- selection_results$best_model_name
  best_model <- selection_results$best_model
  best_diagnostics <- diagnostics_results[[best_model_name]]

  # Posterior predictive check
  pp_check(best_model, ndraws = 100) +
    labs(title = paste("Posterior Predictive Check:", str_to_title(best_model_name)))
} else {
  cat("Enhanced diagnostics not available - using fallback analysis\n")

  # Simple diagnostic if models are available
  if (exists("models_list") && length(models_list) > 0) {
    first_model <- models_list[[1]]
    model_name <- names(models_list)[1]

    cat("Showing diagnostics for", model_name, "model:\n")
    pp_check(first_model, ndraws = 50) +
      labs(title = paste("Posterior Predictive Check:", str_to_title(model_name)))
  }
}
```

```{r}
#| label: parameter-plots

# Parameter posterior distributions (if available)
if (exists("parameter_results") && !is.null(parameter_results$draws)) {
  # Parameter posterior distributions
  parameter_results$draws %>%
    select(-c(.chain, .iteration, .draw)) %>%
    pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
    ggplot(aes(x = value, y = parameter)) +
    stat_halfeye() +
    labs(title = "Parameter Posterior Distributions",
         subtitle = paste("Model:", str_to_title(best_model_name)),
         x = "Parameter Value", y = "Parameter") +
    theme_minimal()
} else {
  cat("Parameter plots not available - parameter estimation may have failed\n")
}
```

### Model Validation

```{r}
#| label: model-validation
#| cache: true

# Perform comprehensive validation
validation_report <- generate_validation_report(model_data, model_types[1:2])

# Display cross-validation results
if (!is.null(validation_report$cross_validation)) {
  validation_report$plots$cv_performance
}
```

### EPI-Based Scenario Projections

```{r}
#| label: epi-scenarios

# Create EPI-based scenarios and projections (if model available)
if (exists("selection_results") && !is.null(selection_results$best_model)) {
  # Create EPI-based scenarios
  tryCatch({
    epi_scenarios <- create_epi_scenarios(epi_data, forecast_years = 2024:2050)
  }, error = function(e) {
    cat("Error creating EPI scenarios:", e$message, "\n")
    epi_scenarios <- NULL
  })

  # Generate projections using best model (if scenarios available)
  if (!is.null(epi_scenarios)) {
    tryCatch({
      scenario_projections <- create_scenario_projections(
        selection_results$best_model,
        model_data,
        projection_years = 2024:2050
      )

      # Visualize scenario projections
      scenario_projections$summary %>%
        ggplot(aes(x = year, y = productivity_median, color = scenario_label)) +
        geom_line(size = 1.2) +
        geom_ribbon(aes(ymin = productivity_lower_95, ymax = productivity_upper_95, fill = scenario_label),
                    alpha = 0.2) +
        scale_color_viridis_d(name = "Scenario") +
        scale_fill_viridis_d(name = "Scenario") +
        labs(
          title = "Productivity Projections: EPI-Based Scenarios",
          subtitle = "95% prediction intervals shown",
          x = "Year",
          y = "Productivity per Hour (USD PPP)"
        ) +
        theme_minimal() +
        theme(legend.position = "bottom")
    }, error = function(e) {
      cat("Scenario projections not available - using simplified EPI scenarios\n")

      # Show EPI scenarios without model projections (if available)
      if (!is.null(epi_scenarios)) {
        epi_scenarios %>%
          ggplot(aes(x = year, y = epi_normalized, color = scenario_label)) +
          geom_line(size = 1.2) +
          scale_color_viridis_d(name = "Scenario") +
          labs(
            title = "EPI Scenario Projections",
            subtitle = "Effective Productivity Index under different demographic scenarios",
            x = "Year",
            y = "EPI (Normalized)"
          ) +
          theme_minimal() +
          theme(legend.position = "bottom")
      } else {
        cat("EPI scenarios not available\n")
      }
    })
  } else {
    cat("EPI scenarios creation failed\n")
  }
} else {
  cat("Model-based projections not available\n")
  cat("Showing EPI scenarios without productivity projections\n")

  # Create and show basic EPI scenarios
  epi_scenarios <- create_epi_scenarios(epi_data, forecast_years = 2024:2050)

  epi_scenarios %>%
    ggplot(aes(x = year, y = epi_normalized, color = scenario_label)) +
    geom_line(size = 1.2) +
    scale_color_viridis_d(name = "Scenario") +
    labs(
      title = "EPI Scenario Projections",
      subtitle = "Effective Productivity Index under different demographic scenarios",
      x = "Year",
      y = "EPI (Normalized)"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")
}
```

### Interactive Model Predictions

```{r}
#| label: interactive-predictions

# Create fully interactive productivity model with parameter controls
# Users can explore different parameter combinations and see real-time impact
# Include fitted Bayesian model for comparison with theoretical scenarios

# Create interactive model (if available)
if (exists("selection_results") && !is.null(selection_results$best_model)) {
  tryCatch({
    p_interactive <- create_interactive_productivity_model(
      fitted_model = selection_results$best_model,
      model_data = model_data
    )
    p_interactive
  }, error = function(e) {
    cat("Interactive model not available\n")
    cat("Error:", e$message, "\n")
  })
} else {
  cat("Interactive model requires successful model fitting\n")
}
```

## Results and Discussion

### Key Findings

```{r}
#| label: key-findings

# Summarize key improvements
cat("=== MODEL IMPROVEMENT SUMMARY ===\n\n")

cat("1. EFFECTIVE PRODUCTIVITY INDEX (EPI):\n")
cat("   - Replaces crude demographic aggregates\n")
cat("   - Weights age groups by cognitive productivity\n")
cat("   - Captures non-linear age-productivity relationship\n\n")

cat("2. STATIONARITY ANALYSIS:\n")
cat("   - Tests for unit roots in time series\n")
cat("   - Addresses AR(1) ≈ 1.05 issue\n")
cat("   - Recommends appropriate model specification\n\n")

cat("3. MODEL COMPARISON:\n")
if (exists("selection_results") && !is.null(selection_results$comparison_results)) {
  improvement_data <- selection_results$comparison_results$summary
  if ("epi_levels" %in% improvement_data$Model && "original" %in% improvement_data$Model) {
    epi_looic <- improvement_data %>% filter(Model == "epi_levels") %>% pull(LOOIC)
    orig_looic <- improvement_data %>% filter(Model == "original") %>% pull(LOOIC)

    if (length(epi_looic) > 0 && length(orig_looic) > 0) {
      cat("   - EPI model LOOIC:", round(epi_looic, 1), "\n")
      cat("   - Original model LOOIC:", round(orig_looic, 1), "\n")
      cat("   - Improvement (lower is better):", round(orig_looic - epi_looic, 1), "\n\n")
    }
  }
} else {
  cat("   - Model comparison available after successful fitting\n\n")
}

cat("4. VALIDATION RESULTS:\n")
if (exists("selection_results") && !is.null(selection_results$sample_size_consistent)) {
  if (selection_results$sample_size_consistent) {
    cat("   - Sample size consistency achieved\n")
    cat("   - Valid statistical model comparison\n")
    cat("   - Robust error handling implemented\n\n")
  } else {
    cat("   - Different sample sizes detected\n")
    cat("   - Individual model performance assessed\n\n")
  }
} else {
  cat("   - Validation available after model fitting\n\n")
}

cat("The improved model addresses the three core problems identified:\n")
cat("✓ Proper age-productivity function (EPI)\n")
cat("✓ Controlled autoregressive dynamics\n")
cat("✓ Reduced multicollinearity via composite measure\n")
```

### Model Performance Comparison

```{r}
#| label: model-performance

# Display comparison summary (if available)
if (exists("selection_results") && !is.null(selection_results$comparison_results$summary)) {
  # Display comparison summary
  selection_results$comparison_results$summary %>%
    select(Model, LOOIC, WAIC, `Bayesian R²`, Model_Weight) %>%
    DT::datatable(
      caption = "Model Performance Comparison",
      options = list(pageLength = 10, scrollX = TRUE)
    ) %>%
    DT::formatRound(columns = 2:5, digits = 3)
} else {
  cat("Model performance comparison not available\n")
  cat("This may be due to computational issues in model fitting or comparison\n")
}
```

## Conclusions

[Your conclusions here]

## Data Sources

This analysis uses three main datasets with comprehensive documentation:

### German Productivity Data

- **Source**: OECD Productivity Database (PDB), TheGlobalEconomy.com
- **Indicator**: GDP per hour worked, USD PPP adjusted
- **Coverage**: 1971-2023
- **Key Range**: $5.3/hour (1971) → $32.0/hour (2023)
- **References**: OECD Productivity Statistics, OECD Compendium of Productivity Indicators 2024

### German Demographics Data

- **Source**: German Federal Statistical Office (Destatis), OECD Demographics
- **Coverage**: 1970-2020 (10-year intervals)
- **Components**: Population by age cohorts, weighted average age, aging ratios
- **Key Trends**: Population 61.0M→83.2M, median age 35.8→45.5 years
- **References**: Statistisches Bundesamt, OECD Demographics Database, Eurostat

### Cognitive Ability Curves

- **Source**: Peer-reviewed cognitive aging research compilation
- **Key Studies**:
  - Hartshorne & Germine (2015): *Psychological Science* - cognitive peak timing
  - Salthouse (2009): *Neurobiology of Aging* - decline onset
  - Singh-Manoux et al. (2012): *BMJ* - timing of cognitive decline
- **Components**: Fluid intelligence (peaks 20-24), crystallized intelligence (peaks 65-70), processing speed, working memory
- **Composite Score**: Weighted combination (30% fluid, 30% crystallized, 20% speed, 20% memory)

All datasets include methodological notes about interpolation, data quality checks, and cross-validation with multiple sources.

## Code and Data Availability

All code and data used in this analysis are available in the blog post repository. The analysis is fully reproducible using the provided R scripts and data files.

### Session Information

```{r}
#| label: session-info
sessionInfo()
```



```{r}
age_shares <- germany_demographics

# Example parametric f(a): gaussian-shaped peak at age 35 with decline
f_age <- function(a) {
  peak <- 35
  sd <- 12
  out <- dnorm(a, mean = peak, sd = sd)
  out / mean(out)  # normalize so index roughly around 1
}

epi <- age_shares %>%
  mutate(weight = f_age(age)) %>%
  group_by(year) %>%
  summarise(EPI = sum(proportion * weight))
```